|

### 2024-11-01 manuscript/paper.Rmd
results text

## Modeling fossil stickleback trait means through time for males and females
 The goal of our analysis was to estimate fossil trait means through time, by sex, while incorporating uncertainty across imputations in the classification of sex. As we explain in this section, we are implementing a complex model structure to account for the fact that the rate of change in the mean of the trait over time may be a non-constant value. We implement a Bayesian approach because these methods are more effective for analyzing the hierarchical model structures we implement. <!--We used a Bayesian approach to estimate trait means in part because imputation is the process of sampling from a posterior predictive distribution of missing data, given observed data. \textcolor{red}{Matt, THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH, EXPLAINING IT TO MY AUDIENCE. -YS}  -->

<!-- \textcolor{red}{MATT, somewhere we need to describe how you handled traits with lots of evolved zeros (ds1, ds2, ds3, lps, tpg) -YS} -->

@@

Continuous traits will likely have some correlation with standard length (i.e., allometry) (@Huxley1932)(@Vojeetal2022)(@Vojeetal2014). We account for this by adding an additional parameter, $\gamma_k$. In addition, the continuous traits ds1, ds2, ds3, lps, tpg are notable because they tend to be lost during the evolution of armor reduction in this system (CITATIONS). This generates meaningful zeros for trait lengths in our empirical dataset that \textcolor{red}{MATT, help describing why we wanted to analyze with and without 0s? -YS}. For these random variables, we added the binary variable $Z_{ij}$, where $Z_{k,ij} = 1$ means that the trait is lost, and $Z_{k,ij} = 0$ means the trait is observed. For all other continuous traits, we assume $Z_{k,ij} = 0$ <!--We also define $Y_{k,ij}$ as -->

@@

If $Y_{k,ij}$ is a continuous trait, then, potentially conditioning on $Z_{k,ij} = 0$, 
\begin{align}
{Y}_{k,ij} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:cont}
\end{align}
In other words, (\ref{eq:cont}) only considers the subset of fossils where the given trait is present when performing a data analysis. For the traits with potential loss, we model the $Z_{k,ij}$ via a $Bernoulli(p_{k,ij})$ distribution where
\begin{align}
\text{logit}({p}_{k,ij}) & = \left\{\begin{array}{llll} \alpha_{k,ft_i} + \beta_kY_{K,ij} & W_{ij} = \text{Female} \\ \alpha_{k,mt_i} + \beta_kY_{K,ij}, & W_{ij} = \text{Male} \end{array}\right.,
\label{eq:cont_prob}
\end{align}
and logit is a function to convert the probability to the log-odds, to allow for easier modelling of linear techniques while still ensuring any probability predictions are guaranteed to be between 0 and 1.

@@

<!-- \textcolor{red}{Q FOR MATT. Is transforming the GP onto the real number line separate from setting the priors? If so, should the statement of transformation go in the section where the GP and discrete trait modeling is discussed. Then priors get their own section here? -YS} -->

@@

$\boldsymbol{\tau}_k = \{\tau_{k,r}: r = 1,\cdots,R\}$ for trait $k$ where $\tau_r$ is the estimate of the Kendall's $\tau$ correlation for $\mu_{k,m,t_i} - \mu_{k,f,t_i}$ versus time $t_i$ for the $r^{th}$ MCMC sample from the posterior distribution. \textcolor{red}{Matt, I think there's an absolute value problem here. Dimorphism can increase of M-F gets more positive or more negative. I've indicated that in the text below, but I could use some help refininig the explanation -YS}
For each trait, the proportion of values in $\boldsymbol{\tau}_k$ that are greater than 0 ($\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)$) may indicate that males have gotten relatively larger than females, indicating more pronounced dimorphism. For each trait, the proportion of values in $\boldsymbol{\tau}_k$ that are less than 0 ($\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} < 0)$) may indicate that females have gotten relatively larger than males, indicating more pronounced dimormphism.

@@

## Sexual Dimorphism 
Table \ref{tab:prob0-table} shows the posterior probability of the differences in the means between male and female specimens at each time point.  Probabilities near 0.5 indicate very little difference in the means whereas probabilities far from 0.5 indicate sexual dimorphism with values of 0 and 1 indicating larger mean values for females and males, respectively.  These results are shown graphically in figure \ref{fig:prob0-plot}.

@@

## Changes over time
We can also assess the strength of the dimorphism by calculating the proportion of posterior samples that are greater than a prespecified threshold \textcolor{red}{(0.5??, idk what this could be.) -MS}

Table \ref{tab:auc-table} shows the probability that the posterior distribution of the differences in the mean is greater than the posterior distribution of the differences at the first time point.  Values near 0.5 indicate very little difference between the distributions whereas values closer to 0 and 1 indicate changes in the distributions with values of 1 meaning the difference in means has shifted towards males having a larger mean and values of 0 indicating the difference in means has shifted towards larger means in females.  These results are shown graphically in figure \ref{fig:AUC-plot}.

>>>>

## Modeling fossil stickleback trait means through time for males and females
 The goal of our analysis was to model fossil trait means through time, by sex, while incorporating uncertainty across imputations in the classification of sex. We therefore implemented a Bayesian hierarchical model structure (described below) to account for classfication uncertainty and the fact that the rate of change of any one trait mean over time may be a non-constant value.

@@

Continuous traits will likely have some correlation with standard length (i.e., allometry) (@Huxley1932)(@Vojeetal2022)(@Vojeetal2014). We account for this by adding an additional parameter, $\gamma_k$. In addition, the continuous traits ds1, ds2, ds3, lps, tpg are notable because they tend to be lost during the evolution of armor reduction in this system (@Belletal1985, @Stuartetal2020). This generates meaningful zeros for trait lengths in our empirical dataset that violate model assumptions. Therefore, for these random variables, we added the binary variable $Z_{ij}$, where $Z_{k,ij} = 1$ means that the trait is lost, and $Z_{k,ij} = 0$ means the trait is observed. For all other continuous traits, we assume $Z_{k,ij} = 0$ <!--We also define $Y_{k,ij}$ as -->

@@

If $Y_{k,ij}$ is a continuous trait, then, potentially conditioning on $Z_{k,ij} = 0$, 
\begin{align}
{Y}_{k,ij} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:cont}
\end{align}
In other words, (\ref{eq:cont}) only considers the subset of fossils where the given trait is present when performing a data analysis. For the traits with potential loss, we model the $Z_{k,ij}$ via a $Bernoulli(p_{k,ij})$ distribution where
\begin{align}
\text{logit}({p}_{k,ij}) & = \left\{\begin{array}{llll} \alpha_{k,ft_i} + \beta_kY_{K,ij} & W_{ij} = \text{Female} \\ \alpha_{k,mt_i} + \beta_kY_{K,ij}, & W_{ij} = \text{Male} \end{array}\right.,
\label{eq:cont_prob}
\end{align}
and logit is a function to convert the probability to the log-odds, to allow for easier modelling of linear techniques while still ensuring any probability predictions are guaranteed to be between 0 and 1. Modeling traits this way allows us to first estimate the probability of trait loss for an individual at a given time, and ask if the sexes differ in their tendency toward loss through time. Then, for individuals that have not lost the trait, we can estimate trait values and calculate sexual dimorphism through time.

@@

@@

$\boldsymbol{\tau}_k = \{\tau_{k,r}: r = 1,\cdots,R\}$ for trait $k$ where $\tau_r$ is the estimate of the Kendall's $\tau$ correlation for $\mu_{k,m,t_i} - \mu_{k,f,t_i}$ versus time $t_i$ for the $r^{th}$ MCMC sample from the posterior distribution. For each trait, a large proportion of values in $\boldsymbol{\tau}_k$ that are greater than 0 ($\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)$) may indicate that males have gotten relatively larger than females through time, indicating an increase in male-biased dimorphism through time. A large proportion of values in $\boldsymbol{\tau}_k$ that are less than 0 ($\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} < 0)$) may indicate that females have gotten relatively larger than males, indicating an increase in female-biased dimorphism through time.

@@

## Sexual Dimorphism

Table \ref{tab:prob0-table} shows the posterior probability of the differences in the means between male and female specimens at each time point.  Probabilities near 0.5 indicate very little difference in the means whereas probabilities far from 0.5 indicate sexual dimorphism with values of 0 and 1 indicating larger mean values for females and males, respectively.  These results are shown graphically in figure \ref{fig:prob0-plot}.

@@

## Changes over time
We estimated the change in sexual dimorphism over time by calculating the distribution of Kendall's Tau for each trait. Here, positive values of Tau indicate that sexual dimorphism (defined as male means minus female means) increased through time. Average Tau values ranged from X to Y across traits and the proportion of positive area under the curve ranged from Z to W (Table Tau). Distributions are available for each trait in the Supplementary Material (Figures SX-SX+15). In Figure (Tau), we show representative distributions. For example,

Something about the probability of dimorphim in the probability of loss? I.e., the prob_zero plots. For example, ds2_prob_zero has lots of positive area under the curve indicating that the rate of loss in males and females diverged, such that males were more likely to lose ds2 through time. Interestingly, for the individuals that kept their spines, dimorphism increased, with females getting larger than males (or males getting smaller than females).

We can also assess the strength of the dimorphism by calculating the proportion of posterior samples that are greater than a prespecified threshold \textcolor{red}{(0.5??, idk what this could be.) -MS}

<!-- Table \ref{tab:auc-table} shows the probability that the posterior distribution of the differences in the mean is greater than the posterior distribution of the differences at the first time point.  Values near 0.5 indicate very little difference between the distributions whereas values closer to 0 and 1 indicate changes in the distributions with values of 1 meaning the difference in means has shifted towards males having a larger mean and values of 0 indicating the difference in means has shifted towards larger means in females.  These results are shown graphically in figure \ref{fig:AUC-plot}. -->
